One-The-Fly Decision Procedures For GKAT

### 2024-12-29 main.tex
big proof done.

With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ ⟨S, ϵ̂, δ̂⟩\) can be lowered into a GKAT coalgebra \(⟨S, δ⟩\) in the following manner:
\[
δ(s, α) ≜ \begin{cases}
    \accept & α ≤ ϵ̂(s) \\  
    (s', p) & ∃ b ∈ \Bool_B, α ≤ b \text{ and } δ(s, b) = (s', p) \\  
    \reject & \text{otherwise}
\end{cases}
\]
This is well-defined, i.e. no more than one clause can be satisfied precisely because the boolean expressions appear in \(ϵ̂\) and \(δ̂\) are disjoint.
The trace semantics of a GKAT coalgebra \(⟨S, ϵ̂, δ̂⟩\) is then defined as the trace semantics of its lowering \(⟨S, δ⟩\).

@@

@@

We will go through two ways of constructing symbolic GKAT automata, one is Thompson's construction, the other is derivative.
However, in order to exhibit the relation between these two construction, specifically, the derivative construction always produces a smaller automaton than the Thompson's construction, we opt to use to produce "weak" symbolic GKAT automata: symbolic GKAT automata where the transition no longer need to perform an action.

\subsection{Weak Symbolic GKAT Automata And Tightening}

Weak Symbolic GKAT automata is a symbolic GKAT automata that allow internal transitions via a boolean expression:
\[⟨S, 
ϵ̂: S → \Bool, 
δ̂: S → (\Bool × S) + (\Bool × S × K)⟩.\]
We also require the deterministic criteria, all the boolean transitions of \(s\), i.e. all the boolean expressions in the following set:
\[\{ϵ̂(s)\} ∪ 
\{b ∣ ∃ s' ∈ S, (b, s') ∈ δ̂(s)\} ∪ 
\{b ∣ ∃ s' ∈ S, p ∈ K, (b, s', p) ∈ δ̂(s)\},\]
are disjoint.
Notice that our notion of

% \section{Symbolic algorithm BAK}

% \subsection{Symbolic GKAT Coalgebra}

% Given an alphabet \(K, B\), a \emph{symbolic GKAT coalgebra} \(Ŝ ≜ ⟨S, ϵ̂, δ̂⟩\) consists of a state set \(S\) consists of a accepting boolean \(ϵ̂\) and a transition function \(δ̂\):
% \begin{mathpar}
%     ϵ̂: S → \Bool_B, \and
%     δ̂: S → S → K → \Bool_B,
% \end{mathpar}
% where \(\Bool_B\) is the free boolean algebra over \(B\) (boolean expressions modulo boolean algebra axioms), and for all states \(s ∈ S\), all the booleans results are ``disjoint''; namely the conjunction of any two expression from the set \(ϵ̂(s) + \{δ̂(s, s', p) ∣ s' ∈ S, p ∈ K\}\) are false in \(\Bool_B\).
% The ordering on two symbolic coalgebra on the same carrier \(S\) is defined pointwise, namely 
% \[⟨ϵ̂₁, δ̂₁⟩ ≤ ⟨ϵ̂₂, δ₂⟩ \text{ if and only if } ∀ s, s' ∈ S, ∀ p ∈ K, \hat{ϵ₁}(s) ≤ \hat{ϵ₂}(s) \text{ and } \hat{δ₁}(s, s', p) ≤ \hat{δ₂}(s, s', p).\]

% Intuitively, the expressions in \(\Bool_B\) is treated as the ``symbolic'' transitions, which can be think of as a set of atoms by the isomorphism \(\Bool_B ≅ 2^{\At_B}\). 
% Then we can explain the intuition for the definition of symbolic GKAT coalgebra:
% \begin{itemize}
%     \item \(ϵ̂(s)\) can be thought of as all the atoms that is accepted by the state \(s\).
%     \item \(δ̂(s, s', p)\) denotes all the atoms that can transition from \(s\) to \(s'\) while executing \(p\).
% \end{itemize}
% Then the atoms excluded by both operations are implicitly rejecting. Formally, we can define a boolean expression that contains all the atoms that will be rejected 
% \[ρ̂(s) = ¬ (ϵ̂(s) ∨ ⋁ \{b ∣ ∃ s' ∈ S, p ∈ K, δ̂(s, s', p) = b\}).\]

% This intuition gives us a isomorphism between between GKAT coalgebra and symbolic GKAT coalgebra. Formally, given a GKAT coalgebra \(⟨S, δ⟩\), then its corresponding symbolic GKAT coalgebra will share the same states \(S\), with the transition defined as follows:
% \begin{align*}
%     ϵ̂(s) & ≜ ⋁ \{α ∈ \At ∣ δ(s, α) = \accept\} & 
%     δ̂(s, s', p) & ⋁ \{α ∈ \At ∣ δ(s, α) = (s', p)\}.
% \end{align*}
% We will denote the above map as \(\symb\), i.e. \(\symb(⟨S, δ⟩) ≜ ⟨S, ϵ̂, δ̂⟩\). Notice that we are implicitly using atoms as an element in the boolean algebra, indeed an atom can be thought about as a conjunction.

% The inverse of \(\symb\) is denoted as \(\unsymb\) where given a symbolic GKAT coalgebra \(⟨S, ϵ̂, δ̂⟩\), we construct a GKAT coalgebra over states \(S\) with the following transition:
% \[δ(s, α) ≜ \begin{cases}
%     \accept & α ∈ ϵ̂(s) \\  
%     (s', p) & α ∈ δ̂(s, s', p) \\  
%     \reject & \text{otherwise}
% \end{cases}\]
% Notice the \(δ\) is well-defined, that is given any \(s ∈ S\) and \(α ∈ \At\), \(δ(s,α)\) can only take on one value; because symbolic GKAT coalgebra is deterministic.

% \subsection{Decision Procedure Of Symbolic GKAT coalgebra}

% The formulation of symbolic GKAT coalgebra is mathematically nice, yet inefficient. Between most states \(s, s' ∈ S\) and action \(p ∈ K\), \(δ(s, s', p)\) will always be false. Therefore, we can developed a set representation of the derivative, where we will typically denote using \(Δ̂\):
% \[Δ̂: S → 𝒫(\Bool_B × S × K).\]
% we always assume transition functions denoted this way are \emph{deterministic}; specifically, all the boolean expression in the following set are disjoint: \[\{b ∣ ∃ s' ∈ S, p ∈ K, (b, s', p) ∈ Δ̂(s)\}.\]

% Intuitively, if \((b, s', p) ∈ Δ̂(s)\), then \(s\) will transition to \(s'\) for all the atoms \(α ≤ b\) while executing \(p\). Notice that the above statement does not imply that \emph{all} \(α\) that makes \(s\) transition to \(s'\) while executing \(p\) will be covered by \(b\); that is, there can exists a \((b', s', p) ∈ Δ̂(s)\), where \(b ≠ b'\).

% This means that the set representation of a symbolic GKAT coalgebra no on longer uniquely correspond to a GKAT coalgebra.
% \begin{example}

% \end{example}

% However, every set representation of a symbolic GKAT coalgebra, indeed correspond to a GKAT coalgebra, specifically given a set representation \(⟨S, ϵ̂, Δ̂⟩\), we can construct a symbolic GKAT \(⟨S, ϵ̂, δ̂⟩\), by defining the following \(δ̂\):
% \[δ̂(s, s', p) = ⋁ \{b ∣ (b, s', p) ∈ Δ̂(s)\}.\]
% More importantly, the map described above is surjective. This means that all the GKAT coalgebra can indeed be represented as at least one set-representation.
% Similarly, set representation can also be converted to GKAT coalgebra: given a set representation \(⟨S, ϵ̂, Δ̂⟩\), we can lower it into a GKAT coalgebra \(⟨S, δ⟩\), by constructing the following \(δ\):
% \[δ(s, α) ≜ \begin{cases}
%     \accept & α ∈ ϵ̂(s) \\  
%     (s', p) & ∃ b ≥ α, (b, s', p) ∈ Δ̂(s) \\  
%     \reject & \text{otherwise}
% \end{cases}\]
% Indeed, lowering the set representation directly to GKAT coalgebra give the same result as converting to symbolic GKAT coalgebra and then lowering. This fact is essential for us to relate these semantics.

% Finally, we can lift the inductive construction theorem (\cref{thm:inductive-construction}) to set representation of GKAT coalgebras, giving a sound and complete symbolic decision procedure. 
% In the following lemmas, we will use \(S, T\) to denote some GKAT coalgebra, and \(Ŝ, T̂\) will be as some set-represented symbolic GKAT coalgebra that corresponds to \(S\). 
% Formally, if we lower \(Ŝ\) to GKAT coalgebra, we will obtain exactly \(S\); similarly for \(T̂\).

% \begin{lemma}
%     We let the \(s ∈ S\) and \(t ∈ T\) be two states, then the following equivalence between conditions in \cref{thm:inductive-construction} hold: 
%     \begin{itemize}
%         \item condition~\labelcref{itm:acc-condition} in inductive construction holds if and only if \(ϵ̂_{Ŝ}(s) = ϵ̂_{T̂}(t)\).
%         \item condition~\labelcref{itm:transition-bisim} holds if and only if for all \((a, s', p) ∈ δ̂_{Ŝ}(s)\) and \((b, t', q) ∈ δ̂_{T̂}(t)\):
%         \[a ∧ b ≠ 0 \text{ and } p = q ⟹ \text{exists a bisimulation \(∼\) between \(S\) and \(T\), } s.t. s' ∼ t'.\]
%         \item condition~\labelcref{itm:transition-dead}, holds if and only if for all \((a, s', p) ∈ δ̂_{Ŝ}(s)\) and \((b, t', q) ∈ δ̂_{T̂}(t)\):
%         \[a ∧ b ≠ 0 \text{ and } p ≠ q ⟹ \text{both \(s\) and \(t\) are dead}.\]
%         \item condition~\labelcref{itm:rej-or-dead} holds if and only if the following holds:
%     \end{itemize}
% \end{lemma}

% \begin{lemma}
%     Given two states \(\)
% \end{lemma}

% \subsection{Symbolic Derivatives and Correctness}

% Given a symbolic GKAT coalgebra \(⟨S, δ̂, ϵ̂⟩\) and a state \(s ∈ S\), we use the notation \(s ⇒ b\) to denote that \(ϵ̂(s) ≥ b\), and we use the notation \(s \transvia{b ∣ p} s'\) to denote that \(δ̂(s, s', p) ≥ b\), where the ordering is the typical ordering in boolean algebra.

% The symbolic derivatives for GKAT forms a DKCT where the states are represented by GKAT expression \(\GKAT_{K, B}\), and the boolean algebra is the free boolean algebra \(\Bool_B\) over the test alphabet \(B\):
% \begin{mathpar}
%     ϵ: \GKAT_{K, B} → \Bool_B; \and  
%     δ: \GKAT_{K, B} → \GKAT_{K, B} → K → \Bool_B.
% \end{mathpar} 
% the accepting map \(ϵ\) and transition map \(δ\) can be defined by the smallest maps that satisfy the following rules:
% \begin{mathpar}
%     \inferrule{\\}{b ⇒ b} \and 
%     \inferrule{\\}{p \transvia{1 ∣ p} 1} \and 
%     \inferrule{e ⇒ a}{e +_b f ⇒ b a} \and 
%     \inferrule{f ⇒ a}{e +_b f ⇒ b̄ a} \and 
%     \inferrule{e \transvia{a ∣ p} e'}{e +_b f \transvia{b ∧ a ∣ p} e'} \and 
%     \inferrule{f \transvia{a ∣ p} f'}{e +_b f \transvia{b̄ ∧ a ∣ p} f'} \and 
%     \inferrule{e ⇒ a \\ f ⇒ b}{e ⋅ f ⇒ a ∧ b} \and 
%     \inferrule{e ⇒ a \\ f \transvia{b ∣ p} f'}{e ⋅ f \transvia{a ∧ b ∣ p} f'} \and 
%     \inferrule{e \transvia{b ∣ p} e'}{e ⋅ f \transvia{b ∣ p} e' ⋅ f} \and 
%     \inferrule{\\}{e^{(b)} ⇒ b̄} \and 
%     \inferrule{e \transvia{a ∣ p} e'}{e^{(b)} \transvia{b ∧ a ∣ p} e' ⋅ {e}^{(b)}}
% \end{mathpar}
% Notice the rules above are very similar to the derivative rules for GKAT~\cite{schmid_GuardedKleeneAlgebra_2021}, this is no coincidence, as we can establish a correspondence of these two to prove correctness of our symbolic algorithm.

% Since the rules listed above is monotonic, that is if we enlarge the boolean in the premise, we also enlarge the boolean in the result. 
% So we can simply pick the largest boolean on the premise, which is either \(ϵ(e)\) or \(δ(e, e', p)\).
% Since the disjunction of two boolean \(a ∨ b\) is the smallest boolean that is greater than both \(a\) and \(b\), thus we can implement \(ϵ\) and \(δ\) by iteration through all the rules, and take the disjunction of all the possible booleans.

% \begin{align*}
%     ϵ(b) & ≜ b 
%         &  ϵ(e +_b f) & ≜ (b ∧ ϵ(e)) ∨ (b̄ ∧ ϵ(f))\\
%     ϵ(q) & ≜ 0
%         & ϵ(e ⋅ f) & ≜ ϵ(e) ∧ ϵ(f)\\  
%     ϵ(e^{(b)}) & ≜ b̄ 
% \end{align*}

% However, the derivative function \[δ: \GKAT_{K, B} → \GKAT_{K, B} → K → \Bool_B,\] poses challenges in the implementation, since \(\GKAT_{K, B}\) is infinite, so it will be impractical to search through all the expressions.
% However, we can treat the function \(δ\) extensionally: \[\GKAT_{K, B} → K → \Bool_B ⊆ 2^{\GKAT_{K, B} × K × \Bool_B},\]
% and then because the boolean expression don't overlap for each input, hence the boolean expression are necessarily unequal; 
% we can model the set by a partial map, thus all the \(δ\) can be represented as a function to the following partial maps.
% \begin{align*}
%     δ & : \GKAT_{K, B} → (\Bool_B ↛ \GKAT_{K, B} × K) \\
%     δ(b) & ≜ \{\} \\
%     δ(q) & ≜ \{1 ↦ (1, p)\}\\  
%     δ(e +_b f) & ≜ \{b ∧ a ↦ (e', p) ∣ a ↦ (e', p) ∈ δ(e)\} 
%         ∪ \{b̄ ∧ a ↦ (f', p) ∣ a ↦ (f', p) ∈ δ(f)\}\\
%     δ(e ⋅ f) & ≜ \{b ↦ (e' ⋅ f, p) ∣ b ↦ (e, p) ∈ δ(e)\}
%         ∪ \{ϵ(e) ∧ b ↦ (f', p) ∣ b ↦ (f', p) ∈ δ(f)\}\\
%     δ(e^{(b)}) & ≜ \{b ∧ a ↦ (e' ⋅ e^{(b)}, p) ∣ a ↦ (e', p) ∈ δ(e)\}.
% \end{align*}
% FIXME: we still need to think about what representation do we want to go with, with this representation, you can also have duplicated K. We need a good example here of why don't we go with map. One good argument is that this representation is complex.

% \subsection{Algorithm and Complexity}

% \section{Implementation and Experiments}

% \subsection{the new norm}

% The theoretical foundation of this algorithm is surprisingly simple, where we only need one minor tweak to the definition of normalization.
% Instead of normalizing both the states \(S\) and the transition function \(δ_S\), we will only normalize the function \(δ_S\) to obtain \(δ_S\).

% Intuitively, this only rejects transitions to dead states, but doesn't remove dead states completely. As we will see later, the bisimulation generated by this notion will be able to contain rejection state, this allows us to only check for liveness after ``discrepancy'' as described in the intro.

% \begin{definition}[normalized homomorphism]
%     A normalized homomorphism \(h: S → T\) is a homomorphism with normalized transition functions:
%     \[
%         \begin{tikzcd}
%             S \ar{r}{h} \ar[swap]{d}{\norm(δ_S)} & T \ar{d}{\norm(δ_T)}\\  
%             G(S) \ar[swap]{r}{G(h)} & G(T)
%         \end{tikzcd}
%     \]
% \end{definition}

% This normalized homomorphism is very close to the definition of homomorphism in normalized GKAT coalgebra, hence inherits many of the properties. 
% Specifically, the language model \(𝒢\) is the final coalgebra in here.

% \subsection{soundness and completeness}

% The definition of normalized bisimulation naturally arise from a weakened form of homomorphism.

% \begin{definition}[Normalizd homomorphism]
%     A \emph{normalized homomorphism} \(h: S → T\) is coproduct (in set) of a \emph{homomorphism} between the normalized coalgebra \(h_{\norm}: \norm(S) → \norm(T)\) and a \emph{function} \(h_{\dead}: \dead{S} → \dead(T)\). Concretely,
%     \[h(s) ≜ \begin{cases}
%         h_{\norm}(s) & \text{if } s ∈ \norm(S) \\  
%         h_{\dead}(s) & \text{if } s ∈ \dead(S)
%     \end{cases}.\]
%     We will sometimes use the coproduct notation \(h = h_{\norm} + h_{\dead}\).
% \end{definition}

% \begin{theorem}
%     GKAT coalgebras under normalized homomorphisms form a category.
% \end{theorem}

% \begin{proof}
%     Recall that the composition operation compose with coproduct:
%     \[\begin{tikzcd}[row sep=3px]
%         {A_1} & {B_1} & {C_1} \\
%         & {+} \\
%         {A_2} & {B_2} & {C_2}
%         \arrow["{x_1}", from=1-1, to=1-2]
%         \arrow["{y_1}", from=1-2, to=1-3]
%         \arrow["{x_2}", from=3-1, to=3-2]
%         \arrow["{y_2}", from=3-2, to=3-3]
%     \end{tikzcd}
%     = 
%     A₁ + A₁ \xrightarrow{x₁ + x₂} B₁ + B₂ \xrightarrow{y₁ + y₂} C₁ + C₂.\]

>>>>

With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ ⟨S, ϵ̂, δ̂⟩\) can be lowered into a GKAT coalgebra \(⟨S, δ⟩\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\  
    (s', p) & ∃ b ∈ \Bool_B, α ≤ b \text{ and } δ(s, b) = (s', p) \\  
    \reject & \text{otherwise}
\end{cases}
\end{align}
This is well-defined, i.e. no more than one clause can be satisfied precisely because the boolean expressions appear in \(ϵ̂\) and \(δ̂\) are disjoint.
The trace semantics of a GKAT coalgebra \(⟨S, ϵ̂, δ̂⟩\) is then defined as the trace semantics of its lowering \(⟨S, δ⟩\).

@@

\begin{theorem}[Functoriality]
    The lowering operation is a functor, given a symbolic GKAT coalgebra homomorphism \(h: Ŝ → Û\), then \(h\) is also a homomorphism \(h: S → U\).
\end{theorem}

\begin{proof}

\end{proof}

@@

The final piece of the puzzle is to convert any given expression into an ``equivalent'' Symbolic GKAT Automata. This goal can be achieved by lifting existent constructions like derivatives and Thompson's construction~\cite{schmid_GuardedKleeneAlgebra_2021,smolka_GuardedKleeneAlgebra_2020}.
The correctness of these conversions is a consequence of  correctness of their non-symbolic counter-part, i.e. we will prove that the lowering as shown in~\labelcref{cons:lowering} of these constructions will yield the conventional derivative and Thompson's construction.

\begin{figure*}
    \begin{mathpar}
        \inferrule[]{\\}
        {p \transvia{1 ∣ p}_{D̂} 1} \and  
        \inferrule[]{\\}
        {b \transAcc{D̂}{b}} \and  
        \inferrule[]
        {e \transvia{c ∣ p}_{D̂} e'}
        {e +_b f \transvia{b ∧ c ∣ p}_{D̂} e'} 
        \and
        \inferrule[]
        {e \transAcc{D̂}{c}}
        {e +_b f \transAcc{D̂}{b ∧ c}}
        \and
        \inferrule[]
        {f \transvia{c ∣ p}_{D̂} f'}
        {e +_b f \transvia{\overline{b} ∧ c ∣ p}_{D̂} f'}
        \and
        \inferrule[]
        {f \transAcc{D̂}{c}}
        {e +_b f \transAcc{D̂}{\overline{b} ∧ c}}
        \and
        \inferrule[]
        {e \transAcc{D̂}{b} \\ f \transAcc{D̂}{c}}
        {e; f \transAcc{D̂}{b ∧ c}}
        \and 
        \inferrule[]
        {e \transvia{b ∣ p}_{D̂} e'}
        {e; f \transvia{b ∣ p}_{D̂} e'; f}
        \and
        \inferrule[]
        {e \transAcc{D̂}{b} \\ f \transvia{c ∣ p}_{D̂} f'}
        {e; f \transvia{b ∧ c ∣ p}_{D̂} f'}
        \and  
        \inferrule[]
        {\\}
        {e^{(b)} \transAcc{D̂}{\overline{b}}}  
        \and  
        \inferrule[]
        {e \transvia{c ∣ p} e'}
        {e^{(b)} \transvia{b ∧ c ∣ p}_{D̂} e'; e^{(b)}}
    \end{mathpar}
    \caption{Symbolic Derivative of GKAT Automata.}\label{fig:derivatives-rules}
\end{figure*}

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the least symbolic GKAT coalgebra (ordered by point-wise subset ordering on \(ϵ̂\) and \(δ̂\)) that satisfy the rules in~\Cref{fig:derivatives-rules}.
Notice that the rules listed on~\Cref{fig:derivatives-rules} is very close to that of Schmid et al.~\cite{schmid_GuardedKleeneAlgebra_2021}.
This is no coincidence, as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis on our correctness argument.

\begin{theorem}[Correctness]
    The lowering of \(D̂\) is exactly the derivative defined by Schmid et. al.~\cite{schmid_GuardedKleeneAlgebra_2021}.
    TODO: unfold the statement.
\end{theorem}

% \begin{table*}
%     \centering
%     \begin{tabular}{c||c|c|c|c|c|c}
%         Exp & \(S\) 
%         & \(ϵ^*\) & \(δ^*\) 
%         & \(ϵ̂(s)\) & \(δ̂(s)\) \\\hline
%         \(b\) & \(∅\) 
%         & \{b\} & \(∅\) 
%         & N/A & N/A \\  
%         \(p\) & \(\{*\}\) 
%         & \(∅\) & \(\{(1, *, p)\}\) 
%         & \(\{1\}\) & \(∅\)\\  
%         \(e₁ +_b e₂\) & \(S₁ + S₂\) &
%         \(⟨\{b\}|~ϵ₁^* ∪ ⟨\{\overline{b}\}|~ϵ₂^*\) &
%         \(⟨\{b\}|~δ₁^* + ⟨\{\overline{b}\}|~δ₂^*\) & 
%         \(\begin{cases}
%             ϵ̂₁(s) & s ∈ S₁\\
%             ϵ̂₂(s) & s ∈ S₂\\
%         \end{cases}\) & 
%         \(\begin{cases}
%             δ̂₁(s) & s ∈ S₁\\
%             δ̂₂(s) & s ∈ S₂\\
%         \end{cases}\) \\  
%         \(e₁ ; e₂\) & \(S₁ + S₂\) & 
%         \(⟨ϵ₁^*|~ϵ₂^*\) & \(δ₁^* + ⟨ϵ₁^*|~δ₂^*\) & 
%         \(\begin{cases}
%             ⟨ϵ̂₁(s)| ~ ϵ₂^*& s ∈ S₁ \\  
%             ϵ̂₂(s) & s ∈ S₂
%         \end{cases}\)& 
%         \(\begin{cases}
%             δ̂₁(s) ∪ ⟨ϵ̂(s)|~δ₂^* & s ∈ S₁ \\  
%             δ̂2(s) & s ∈ S₂
%         \end{cases}\) \\  
%         \(e₁^{(b)}\) & \(S₁\) & 
%         \(\{\overline{b}\}\) & \(⟨\{b\}|~δ₁^*\) & 
%         \(⟨\{\overline{b}\}|~ϵ̂1(s)\) & 
%         \( ⟨\{b\}|⟨ϵ̂1(s)|~δ₁^* ∪ δ₁(s)\)
%     \end{tabular}
%     \caption{Symbolic Thompson's Construction}\label{tab:symb-thomposon-construction}
% \end{table*}

\begin{table*}
    \centering
    \begin{tabular}{c||c|c|l|l}
        Exp & \(S\) & \(s^*\)  
        & \(ϵ̂(s)\) & \(δ̂(s)\) \\\hline
        \(b\) & \(\{s^*\}\) 
        & \(s^*\)
        & \(\{b\}\) & \(∅\) \\  
        \(p\) & \(\{s^*, s₁\}\) 
        & \(s^*\)
        & \(\begin{cases}
           ∅ & s = s^* \\  
           \{1\} & s = s₁ 
        \end{cases}\) 
        & \(\begin{cases}
            \{(1, s₁, 0)\} & s = s^* \\  
            ∅ & s = s₁
        \end{cases}\)\\  
        \(e₁ +_b e₂\) & \(\{s^*\} + S₁ + S₂\) &
        \(s^*\) &
        \(\begin{cases}
            ⟨\{b\}| ϵ̂₁(s₁^*) ∪ ⟨\{b\}| ϵ̂₂(s₂^*) & s = s^* \\
            ϵ̂₁(s) & s ∈ S₁\\
            ϵ̂₂(s) & s ∈ S₂\\
        \end{cases}\) & 
        \(\begin{cases}
            ⟨\{b\}| δ̂₁(s₁^*) + ⟨\{b\}| δ̂₂(s₂^*) & s = s^* \\
            δ̂₁(s) & s ∈ S₁\\
            δ̂₂(s) & s ∈ S₂\\
        \end{cases}\) \\  
        \(e₁ ; e₂\) & \(S₁ + S₂\) & 
        \(s₁^*\) & 
        \(\begin{cases}
            ⟨ϵ̂₁(s)| ϵ̂₂(s₂^*)& s ∈ S₁ \\  
            ϵ̂₂(s) & s ∈ S₂
        \end{cases}\)& 
        \(\begin{cases}
            δ̂₁(s) + ⟨ϵ̂(s)| δ̂₂(s₂^*) & s ∈ S₁ \\  
            δ̂2(s) & s ∈ S₂
        \end{cases}\) \\  
        \(e₁^{(b)}\) & \(\{s^*\} + S₁\) & 
        \(s^*\) &
        \(\begin{cases}
            \{\overline{b}\} & s = s^*\\
            ⟨\{\overline{b}\}| ϵ̂1(s) & s ∈ S₁
        \end{cases}\) & 
        \(\begin{cases}
            ⟨\{b\}| δ̂₁(s₁^*) & s = s^* \\
            δ₁(s) ∪ ⟨\{b\}|⟨ϵ̂1(s)| δ̂₁(s₁^*) & s ∈ S₁
        \end{cases}\)
    \end{tabular}
    \caption{Symbolic Thompson's Construction}\label{tab:symb-thomposon-construction}
\end{table*}

Another way to construct an automaton is via Thompson's construction, we lift the original construction to the symbolic setting.
A common expression to construct is a guard operation, denoted by \(⟨B|\), where \(B\) is a set of boolean expressions.
TODO: define transition dynamics and accepting dynamics earlier.
Concretely, this guard can be defined on both accepting dynamics and transition dynamics:
\begin{align*}
    ⟨B|ϵ̂(s) & ≜ \{b ∧ c ∣ b ∈ B, c ∈ ϵ(s)\}; \\
    ⟨B|δ̂(s) & ≜ \{(b ∧ c, s', p) ∣ b ∈ B, (c, s', p) ∈ δ(s)\}.
\end{align*}
Notably, besides guarding transition and acceptance with different conditions, like in if statements, the guard expression can be used to simulate uniform continuation. Specifically, we can use \(⟨ϵ̂(s)|~δ(s)\) to connecting all the accepting state of \(s\) to the dynamic \(δ(s)\).

With these definitions in mind, we can define symbolic Thompson's construction inductively as in~\Cref{tab:symb-thomposon-construction}, where we let \((S₁, ϵ̂₁, δ̂₁)\) and \((S₂, ϵ̂₂, δ̂₂)\) to be result of Thompson's construction for \(e₁\) and \(e₂\) respectively.
The \(S₁ + S₂\) is the disjoint union of \(S₁\) and \(S₂\), and for any two transition dynamics \(δ₁(s₁): 𝒫(\Bool × S₁ × K)\) and \(δ₂(s₂): 𝒫(\Bool × S₂ × K)\), then \(δ₁^*(s₁) + δ₂^*(s₂): 𝒫(\Bool × (S₁ + S₂) × K)\) is the bifunctorial lift via \(+\).

One notable difference between the original construction~\cite{smolka_GuardedKleeneAlgebra_2020} and our construction is that we use a start state \(s^* ∈ S\), instead of a start dynamics (or pseudo-state).
This choice will make the proof slightly easier. 
However, in~\Cref{sec:optimization-implementation}, we will explain that our implementation uses start dynamics instead of start state, to avoid unnecessary lookups and unreachable states.

We would like to explore several desirable theoretical properties of both derivatives and Thompson's construction.
Specifically, the correctness, i.e. the semantics of the ``start state'' the both construction have the same preserves the trace semantics of the expression; finiteness, i.e. the coalgebra generated is always finite, which means that our equivalence algorithm will eventually terminate; and finally, how does the number of reachable state relate to the size of the expression, so that we can estimate the complexity of the equivalence checking algorithm.
Turns out all of these questions can be answered by a connection by a homomorphism from symbolic Thompson's construction to the symbolic derivatives.

\begin{theorem}\label{thm:hom-thompson-derivative}
    Given any GKAT expression \(e\), the resulting symbolic GKAT coalgebra from Thompson's construction \(Ŝ_e\) have a homomorphism to derivatives \(h: Ŝ_e → D̂\), s.t. for the start state \(s^* ∈ S, h(s^*) = e\).
\end{theorem}

\begin{proof}
    By induction on the structure of \(e\). We will recall that \(h: Ŝ_e → ⟨e⟩_D\) is a symbolic GKAT coalgebra homomorphism when the following two conditions are true: \(s \transAcc{Sₑ}{b}\) if and only if \(h(s) \transAcc{D̂}{b}\); and \(s \transvia{b ∣ p}_{Sₑ} s'\) if and only if \(h(s) \transvia{b ∣ p}_{D̂} h(s')\).

    When \(e ≜ b\) for some tests \(b\), then the function \(h\) is defined as \(\{s^* ↦ b\}\).
    When \(e ≜ p\) for some primitive action \(p\), then the function \(h\) is defined as \(\{s^* ↦ p, * ↦ 1\}\).
    The homomorphism condition can then be verified by unfolding the definition.

    When \(e ≜ e₁ +_b e₂\), by induction hypothesis, we have homomorphisms \(h₁: Ŝ_{e₁} → ⟨e₁⟩_D\) and \(h₂: Ŝ_{e₂} → ⟨e₂⟩_D\).
    Then we define the homomorphism 
    \[h(s) ≜ \begin{cases}
        e₁ +_b e₂ & s = s^* \\  
        h₁(s) & s ∈ Ŝ_{e₁} \\  
        h₂(s) & s ∈ Ŝ_{e₂}
    \end{cases}\]
    We show that \(h\) is a homomorphism. Because \(Ŝₑ\) preserves the transition and acceptance of \(Ŝ_{e₁}\) and \(Ŝ_{e₂}\), then for all \(s ∈ Ŝ_{e₁} ∩ Ŝ_{e}\), we have
    \begin{align*}
        & s \transAcc{Ŝ_{e}}{c} \text{ iff }
        s \transAcc{Ŝ_{e₁}}{c} \text{ iff }
        h₁(s) \transAcc{D̂}{c} \text{ iff }
        h(s) \transAcc{D̂}{c} \\ 
        & s \transvia{c ∣ p}_{Ŝ_{e}} s' \text{ iff }
        s \transvia{c ∣ p}_{Ŝ_{e₁}} s' \text{ iff }
        h₁(s) \transvia{c ∣ p}_{D̂} h₁(s') \text{ iff }
        h(s) \transvia{c ∣ p}_{D̂} h(s')  
    \end{align*}
    And similarly for \(s ∈ Ŝ_{e₂} ∩ Ŝ_{e}\).
    So we only need to show the homomorphic condition for the start state \(s^*\):
    \begin{align*}
        & s^* \transAcc{Ŝ_{e}}{c} \\
        \text{iff } &
        (∃ a, b ∧ a = c \text{ and } 
        s₁^* \transAcc{Ŝ_{e₁}}{a})
        \text{ or } 
        (∃ a, \overline{b} ∧ a = c \text{ and } 
        s₂^* \transAcc{Ŝ_{e₂}}{a}) \\
        \text{iff } & 
        (∃ a, b ∧ a = c \text{ and } 
        h₁(s₁^*) \transAcc{D̂}{a})
        \text{ or } 
        (∃ a, \overline{b} ∧ a = c \text{ and } 
        h₂(s₂^*) \transAcc{D̂}{a}) \\  
        \text{iff } & 
        (∃ a, b ∧ a = c \text{ and } 
        e₁ \transAcc{D̂}{a})
        \text{ or } 
        (∃ a, \overline{b} ∧ a = c \text{ and } 
        e₂ \transAcc{D̂}{a}) \\  
        \text{iff }& e₁ +_b e₂ \transAcc{D̂}{c}\\
        \text{iff }& h(s^*) \transAcc{D̂}{c}. \\[5px]
        & s^* \transvia{a ∣ p}_{Ŝ_{e}} s' \\
        \text{iff } &
        (∃ a, b ∧ a = c \text{ and } 
        s₁^* \transvia{a ∣ p}_{Ŝ_{e₁}} s')
        \text{ or } 
        (∃ a, \overline{b} ∧ a = c \text{ and } 
        s₂^* \transvia{a ∣ p}_{Ŝ_{e₂}} s') \\
        \text{iff } & 
        (∃ a, b ∧ a = c \text{ and } 
        h₁(s₁^*) \transvia{a ∣ p}_{D̂} h(s'))
        \text{ or } 
        (∃ a, \overline{b} ∧ a = c \text{ and } 
        h₂(s₂^*) \transvia{a ∣ p}_{D̂} h(s')) \\  
        \text{iff } & 
        (∃ a, b ∧ a = c \text{ and } 
        e₁ \transvia{a ∣ p}_{D̂} h(s'))
        \text{ or } 
        (∃ a, \overline{b} ∧ a = c \text{ and } 
        e₂ \transvia{a ∣ p}_{D̂} h(s')) \\  
        \text{iff }& e₁ +_b e₂ \transvia{a ∣ p}_{D̂} h(s') \\ 
        \text{iff }& h(s^*) \transvia{a ∣ p}_{D̂} h(s')
    \end{align*}
